---
title: "Announcing the 'emstreeR' package"
description: |
  An R package for Fast Computing Euclidean Minimum Spanning Trees.
  Originally posted on https://allanvc.github.io
creative_commons: CC BY
author:
  - name: Allan V. C. Quadros
    url: https://allanvc.github.io
date: 12-25-2018
output:
  radix::radix_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

I am proud to announce my first **R** package: **emstreeR**. It is available on CRAN since the beginning of December 2018.

'emstreeR' is a package for fast and easily computing Euclidean Minimum Spanning Trees. It uses 'mlpack' - the C++ Machine Learning library, which is called on background. 'emstreeR' works as a wrapper, so that R users can benefit from the C++ function for computing an Euclidean Minimum Spanning Tree without actually touching the C++ code. This is only possible because 'emstreeR' heavily relies on RcppMLPACK and Rcpp. The package also provides functions and an S3 method for readily plotting the Minimum Spanning Trees (MST) using either 'base', 'scatterplot3d' or 'ggplot2' style.

Working with 'emstreeR' is easy because you just have to pass your matrix or data.frame of points to the main function `computeMST()` in order to get things working.

But, before we begin, what exactly is a Minimum Spanning Tree?

In order to explain that, I will try not to bring any graph concept to the discussion.

Imagine you have to connect some South American cities cities with pipelines in order to distribute the crude oil produced in an oil platform in the coast of Rio de Janeiro. To optimize your operations, you need to know the most efficient route, e.g. the route that connects all the ten cities within the minimum distance as possible. That optimal route between the cities is a Minimum Spanning Tree. Therefore, the Minimum Spanning Tree problem is, in its essence, an optimization problem. 

The distances between the points can be calculated using several methods, such as Manhattan, Euclidean, Minkowski, etc. In 'emstreeR', we use Euclidean distances. Hence, the name Euclidean Minimum Spanning Trees.

So, let's try to reproduce the aforementioned example. 

First, we are going to use the package 'ggmap'. We will pass the cities names and then get their latitude and longitude coordinates.

```{r, echo=TRUE, message=FALSE}
library(ggmap)

## cities:
cities_location <- data.frame(location = c("Passo Fundo, Brazil", 
                                            "Buenos Aires, Argentina",
                                            "Assuncion, Paraguay",
                                            "Campinas, Brazil",
                                            "Cubatao, Brazil",
                                            "Mendoza, Argentina",
                                            "Corrientes, Argentina",
                                            "Porto Velho, Brazil",
                                            "Manaus, Brazil",
                                            "Santa Cruz de La Sierra, Bolivia",
                                            "Belo Horizonte, Brazil",
                                            "Brasília, Brazil",
                                            "Rio de Janeiro, Brazil",
                                            "Campo Grande, Brazil",
                                            "Recife, Brazil"
                                            ), 
                                stringsAsFactors = FALSE)
    
## get latitude and longitude
geo_location <- geocode(cities_location$location, source = "dsk")
    
## combine data:
df_location <- cbind(cities_location, geo_location)
```

Let's plot the map, so we can have an initial notion about what we are dealing with.

```{r, preview=FALSE, echo=TRUE, fig.height=4, fig.width=8, message=FALSE}
# coordinates for South America
map_grid <- c(left = -80, bottom = -40, right = -20, top = 0) 

library(dplyr) # we wanna use pipes

get_stamenmap(map_grid, zoom = 5) %>% ggmap()+
  geom_point(data = df_location,
           aes(x = lon, y = lat), size=2)
  
```


Now is when **emstreeR** comes in handy. We pass the point columns from the df_location object and it computes the minimum route connecting all the points, e.g. the Euclidean minimum spanning tree.


```{r,  echo=TRUE, message=FALSE}
## MST:
library(emstreeR)
out <- ComputeMST(df_location[,2:3])

out
```

As a result, we have a data.frame containing our original data and three more columns: from, to, and distance. This information is the Minimum Spanning Tree and will be used for plotting.

You can silence the output of the algorithm by setting `verbose=FALSE` in `computeMST()`.

To better visualize the result, we plot the map again, but now using the ggplot2's extension provided in 'emstreeR': `stat_MST()`.

```{r, preview=TRUE, fig.height=5, fig.width=8, echo=TRUE, message=FALSE}
## Plot:
# we use the same grid map    
get_stamenmap(map_grid, zoom = 5) %>% ggmap()+
  stat_MST(data = out,
           aes(x = lon, y = lat, from=from, to=to), 
           colour="red", linetype = 2)+
  geom_point(data = out, aes(x = lon, y = lat), size=3)
  
```

Once you have the information about the minimum route, it is easy to see that you will have to install oil distribution hubs in cities that will have more than one leaving or arriving link, such as Assuncion in Paraguay. Based on that information or other criteria such as connectivity or even by cutting the largest edge, you can also define or separate regions of operations. What I mean by this is that you can also use Minimum Spanning Trees in clustering problems. Nonetheless, let's save that topic for another post.

Below, I will present other things you can do with 'emstreeR'.

'emstreeR' provides an S3 method for the 'base' function `plot()`. It is easy to make simple 2D plots with that.

```{r, echo=TRUE, message=FALSE}
## artifical data for 2D plots:
set.seed(1984)
n <- 15
c1 <- data.frame(x = rnorm(n,-0.2, sd=0.2), y = rnorm(n,-2,sd=0.2))
c2 <- data.frame(x = rnorm(n,-1.1, sd=0.15), y = rnorm(n,-2,sd=0.3)) 
d <- rbind(c1, c2)
d <- as.data.frame(d)
  
## MST:
#library(emstreeR)
out <- ComputeMST(d, verbose = FALSE)  
```

```{r, preview=FALSE, fig.height=5, fig.width=7, echo=TRUE, message=FALSE}
## simple 2D plot:
plot(out, col.pts = "red", col.segts = "blue")
```

The `stat_MST()` extension also makes it easy plotting 2D plots, but in ggplot2 style:

```{r, preview=FALSE, fig.height=5, fig.width=7, echo=TRUE, message=FALSE}
## 2D plot with ggplot2:
library(ggplot2)
ggplot(data = out, aes(x = x, y = y, from = from, to = to))+ 
  geom_point()+ 
  stat_MST(colour="red")
```

If you want something fancier, you can use the `geom=curve` argument:

```{r, preview=FALSE, fig.height=5, fig.width=7, echo=TRUE, message=FALSE}
## 2D plot with ggplot2:
#library(ggplot2)
ggplot(data = out, aes(x = x, y = y, from=from, to=to))+ 
  geom_point()+ 
  stat_MST(geom="curve")
```

You can also make 3D MST plots with 'emstreeR'. It provides a wrapper for plotting 3D scatter plots, using the 'scatterplot3d' package.

Let's create 3D artificial data and compute the MST.

```{r, echo=TRUE, message=FALSE}
## artificial data for 3D plots:
n = 99
set.seed(1984)
d1<-matrix(rnorm(n,mean=-2,sd=.5), n/3, 3) # 3d
d2<-matrix(rnorm(n,mean=0,sd=.3), n/3, 3)
d3<-matrix(rnorm(n,mean=3,sd=.4), n/3, 3)
d<-rbind(d1,d2,d3) # just to show a matrix input
  
## MST:
library(emstreeR)
out <- ComputeMST(d, verbose = FALSE)
```

We can do a simple 3D plot using the function `plotMST3D()`.

```{r, preview=FALSE, fig.height=5, fig.width=7, echo=TRUE, message=FALSE}
## simple 3D plot:
plotMST3D(out, xlab = "xaxis", main="Just a MST 3D plot")
```

But, again, if you like fancier plots, we can also make a 3D MST plot using 'plotly':

```{r, preview=FALSE, fig.height=4, fig.width=8, echo=TRUE, message=FALSE}

# some adjusts needed
ord_id <- Reduce(rbind, rbind(t(out[, c("from", "to")]), NA))

ord_data <- out[ord_id, c("X1", "X2", "X3")]
ord_data <- as.data.frame(ord_data)


library(dplyr) # we wanna use pipes
library(plotly)

plot_ly(showlegend = T) %>%
  add_markers(data=ord_data, x = ~X1, y = ~X2, z = ~X3, 
              marker=list(opacity=0.5), size = I(3.5), name="points") %>%
  add_paths(data=ord_data, x = ~X1, y = ~X2, z = ~X3, 
            color = I("red"), name="edges")
```


And that's it! I hope you enjoy **emstreeR**.

* CRAN version: https://cran.r-project.org/package=emstreeR
* Dev version: https://github.com/allanvc/emstreeR
